On the number of tiles visited by a\\ line segment on a rectangular grid

### 2023-04-10 paper.tex
Redefined g
g_new(s,u,v) equals g_old(s,2*u, 2*v)
Fixing issue found by reviewer

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing than \mid
\newcommand{\diff}{\,\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
%\newcommand{\funca}{\psi} % function that gives average number of grid line crossings for real-valued length
\newcommand{\probmax}{\rho} % function that gives the probability of achieving the maximum number of touched tiles in the square case for real-valued length
\newcommand{\ras}{\sigma}
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genfun}{q}
\newcommand{\genvar}{s}
\newcommand{\roundterm}{r}
\newcommand{\mss}{M}

@@

\begin{abstract}
Consider a line segment placed on a two-dimensional grid of rectangular tiles. This paper addresses the relationship between the length of the segment and the number of tiles it visits (i.e. has intersection with). The square grid is also considered explicitly, as some of the specific problems studied are more tractable in that particular case. The segment position and orientation can be modelled as either deterministic or random. In the deterministic setting, the maximum possible number of visited tiles is characterized for a given length, and conversely, the infimum segment length needed to visit a desired number of tiles is analyzed. In the random setting, the average number of visited tiles and the probability of visiting the maximum number of tiles on a square grid are studied as a function of segment length. These questions are related to Buffon's needle problem and its extension by Laplace.

@@

As an example of question \ref{prob: determ dir}, consider $a=1.35$, $b=1$. A segment of unit length can be placed as shown in Figure~\ref{fig: examples} (left) to make it visit $3$ tiles. In fact, this is the maximum number for $\len=1$. The figure also illustrates that the solution for length $2.4$ is $5$ (center), and for $4.7$ it is $8$ (right).

@@

The \emph{inverse} problem \ref{prob: determ inv} is, given $\tiles \in \mathbb N$, to determine the infimum length of all segments that visit at least $\tiles$ tiles. If the length can take any real value the infimum is not a minimum, because given any segment it can be shortened by some small amount without changing the number of visited tiles. This is a consequence of the interior of each tile being an open set.

@@

To address the remaining two questions, the notion of a \emph{random} segment of a given length needs to be precisely defined. This is done as follows. By symmetry, one endpoint of the segment can be assumed to lie in a fixed, reference tile. The position of this endpoint is \emph{uniformly} distributed on the tile. The segment orientation has a \emph{uniform} distribution on $[0,2\pi)$, and is \emph{independent} of the endpoint position. Solving the problem \ref{prob: rand ave} of how many tiles the segment visits on average also answers, as will be seen, the inverse question (segment length to visit a given number of tiles on average). A natural, related problem \ref{prob: rand prob} is with what probability the segment visits the maximum number of tiles.

@@

The Buffon-Laplace needle problem \cite[section~1.1]{Mathai99} considers a needle randomly placed on a grid of rectangular tiles of width $a$ and height $b$. The number of visited tiles equals one plus the number of crossings almost surely. The probability of the needle staying within a single tile is computed in \cite{Arnow94} for the case where $\len < \min\{a,b\}$. One of the problems considered in this paper, as mentioned above, is the complementary question \ref{prob: rand prob} of the probability that the needle visits the maximum number of tiles possible for its length.

@@

For arbitrary $a, b \in \mathbb R^+$, the functions $\funt$ and $\funl$ can be computed using an iterative procedure, which exploits the fact that the pairs $(i_3,j_3), (i_4,j_4), \ldots$ of the set $\mss$ are sorted by increasing $i+j-1$, and also by increasing $(i-2)^2 a^2 + (j-2)^2 b^2$. Namely, for $\funt$ the following method yields the solution: generate successive pairs to find the last one, $(i_\tiles,j_\tiles)$, that satisfies \eqref{eq: len > sqrt}; then $\funt(\len) = \tiles$. For $\funl$ the analogous method gives a direct formula. In addition, it is possible to obtain a direct formula also for $\funt$ using a different approach. These formulas are given in Theorems \ref{theo: funt, form} and \ref{theo: funl, form}.

@@

Theorems \ref{theo: funt, form} and \ref{theo: funl, form} not only give the solutions $\funt(\len)$ and $\funl(\tiles)$ to the first two questions posed in \S\ref{part: intro}; they also provide a way to actually position a segment of length $\len$ or slightly greater than $\funl(\tiles)$, respectively, so that it visits $\funt(\len)$ or $\tiles$ tiles. Namely, for $\isoli$, $\jsoli$ computed as in the corresponding theorem, the segment should have its endpoints in the interiors of two tiles shifted $\isoli-1$ steps horizontally and $\jsoli-1$ steps vertically with respect to each other, with the exact position and orientation of the segment adjusted to avoid any grid points.

@@

For $0 \leq \genvar \leq 1$, let
\begin{equation}
\label{eq: f}
f(\genvar) = \int_0^\genvar \arccos z \diff z = \genvar \arccos \genvar - \sqrt{1-\genvar^2} + 1.
\end{equation}

@@

In the first two cases above, the probability that $x_2 > a(n-1)$, conditioned on $x_1$, is the length of the arc to the right of the line $x=a(n-1)$ divided by $2\pi\len$, that is,
\begin{equation}
\label{eq: Pr x_2 > a(n-1) cond}
\Pr[x_2 > a(n-1) \cond x_1] = \frac 1 \pi \arccos \frac{a(n-1)-x_1}{\len}.
\end{equation}
In the first case $x_1$ has a uniform distribution on $[0,a)$, and $\Pr[x_2 > a(n-1)]$ is easily obtained from \eqref{eq: Pr x_2 > a(n-1) cond} as
\begin{equation}
\label{eq: Pr x_2 > a(n-1), first case}
\begin{split}
\Pr[x_2 > a(n-1)] &= \E[\Pr[x_2 > a(n-1) \cond x_1]] \\
&= \frac 1 {\pi a} \int_0^a\arccos \frac{a(n-1)-x_1}{\len} \diff x_1 \\
&= \frac{\len}{\pi a}\,\, \left( f\left( \frac{a(n-1)}{\len} \right) - f\left( \frac{a(n-2)}{\len} \right) \right),
\end{split}
\end{equation}
where the function $f$ is defined in \eqref{eq: f}. Substituting into \eqref{eq: Pr i = 2 Pr} yields the result in \eqref{eq: Pr i geq n}, second line.

@@

For $u, v \in \mathbb N$, $\genvar \in \mathbb R^+$, let
\begin{equation}
\label{eq: g def}
g(\genvar, u, v) = \frac 2 \pi \int_{\arcsin \frac u {2\genvar}}^{\arccos \frac v {2\genvar}} \left( \genvar \sin \theta - \frac u 2 \right) \left( \genvar \cos \theta - \frac v 2 \right) \diff \theta.
\end{equation}
Computing this integral and using the identities $\cos^2 \alpha = 1-\sin^2 \alpha$ and $\cos(2\alpha) = 2\cos^2\alpha-1 = 1 - 2\sin^2\alpha$,
\begin{equation}
\label{eq: g fin}
\begin{split}
g(\genvar, u, v) &= \frac 1 {2\pi} \biggl(
\left(\arccos\frac{v}{2\genvar}-\arcsin\frac{u}{2\genvar}\right) u v + 2\genvar^2 + \frac{u^2} 2 + \frac{v^2} 2 \\
& \quad  - u \sqrt{4\genvar^2-v^2} - v \sqrt{4\genvar^2-u^2} \biggr).
\end{split}
\end{equation}
Taking into account that $\arcsin x = \pi/2 - \arccos x$, it follows from \eqref{eq: g fin} that $g(\genvar, u, v) = g(\genvar, v, u)$.

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles is
\begin{equation}
\label{eq: probmax}
\begin{split}
\probmax(\len) & = g\left(\len, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } \tiles \text{ odd, } \textstyle \len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right] \\
\probmax(\len) & = g\left(\len, \tiles-3, \tiles-3 \right) + 2 g\left(\len, \tiles-5, \tiles-1 \right) \\
& \text{for } \tiles \text{ odd, } \textstyle \len \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right] \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) \quad \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right] \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) + 2 g\left(\len, \tiles-6, \tiles \right) \\
& \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

@@

For $i,j \in \mathbb Z$, let the (square) tile with lower-left coordinates $(i,j)$ be defined as $[i,i+1) \times [j,j+1)$, i.e.~including its lower and left edges. The first endpoint of the segment is contained in the tile with lower-left corner $(0,0)$, which will be referred to as reference tile. For an arbitrary value $\tiles$, the segment visits $\tiles$ tiles almost surely if its second endpoint is in a tile with lower-left corner $(i-1,j-1)$ such that $i,j\geq 1$ and $i+j-1=\tiles$, or in a tile symmetrical to this about the line $x=1/2$ (with $i \leq 0$), or about $y=1/2$ (with $j \leq 0$), or about both.

Given $\len \in \mathbb R^+$, let $\tiles = \funt(\len)$, and consider $i,j\geq 1$ such that $i+j-1=\tiles$. For a fixed orientation $\theta$, the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ if and only if shifting the segment so that its second endpoint has coordinates $(i-1,j-1)$ results in the first endpoint being in the reference tile. It cannot be the case that the segment ``overshoots'' past this tile, because that would result in a number of visited tiles exceeding $\tiles$, which is the maximum.

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-1)/2,(\tiles-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-3)/2,(\tiles+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions for a given $\tiles = \funt(\len)$ on a unit square grid, with $\tiles$ odd. Example for $\tiles = 7$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

There is a range of possible values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For convenience, the angle $\theta$ is defined clockwise from the downward direction. For each $\theta$ in the allowed range, the valid positions for the first endpoint of the (shifted) segment are in a rectangle contained in the reference tile. The rectangle associated with an example $\theta$ is shown in the figure, shaded. The area of this rectangle is the probability $\probmax_{i-1,j-1}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. The desired probability $\probmax(\len)$ will be obtained by averaging $\probmax_{i-1,j-1}(\len, \theta)$ with respect $\theta$, summing over all $i, j \geq 1$ such that the tile with lower-left corner $(i-1, j-1)$ can be reached from the reference tile, and then using symmetry to take into account the tiles with $i \leq 0$ or $j \leq 0$.

The situation is different depending on whether $\tiles = \funt(\len)$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths for which $\funt(\len)$ equals $\tiles$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference tile from which the tile with lower-left coordinates $((\tiles-1)/2, (\tiles-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((\tiles-1)/2, (\tiles-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((\tiles-3)/2, (\tiles+1)/2)$ and $((\tiles+1)/2, (\tiles-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). By symmetry, each of those two tiles contributes the same probability. Other tiles with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, $i,j\geq 1$, $i+j-1 = \tiles$ do not need to be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

For $\tiles$ even, which corresponds to lengths in the interval \eqref{eq: square, even max tiles, lengths}, the two tiles with lower-left coordinates $(\tiles/2-1, \tiles/2)$ and $(\tiles/2, \tiles/2-1)$ can always be reached from some positions within the reference tile. In addition, if
\begin{equation}
\label{eq: probmax len even2}
\len > \frac{\sqrt{((\tiles-6)^2 + \tiles^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $(\tiles/2-2, \tiles/2+1)$ and $(\tiles/2+1, \tiles/2-2)$ have to be considered too. This can only happen for $\tiles \geq 8$. No other tiles with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, $i+j-1 = \tiles$ can be reached from the reference tile.

From the above two paragraphs it stems that, among all possible tiles with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, those whose centers lie on the lines $x=1/2$ or $y=1/2$ (i.e.~those with $i=1$ of $j=1$) are never reachable from the reference tile, in the sense that a segment that touches the maximum number of tiles cannot have its second endpoint in those tiles. This implies that $\probmax(\len)$ can be computed by considering only the tiles with $i,j\geq 1$ (actually $i,j \geq 2$), and then multiplying by $4$ to include the symmetrical tiles with respect to the lines $x=1/2$ and $y=1/2$. As an example, Figure~\ref{fig: probmax_possible_tiles} shows the reference tile (solid line) and the tiles that can contain the second endpoint (dotted lines) for $\funt(\len) = 5$.

\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Possible tiles for the endpoints of a segment that visits the maximum number of tiles $\tiles= \funt(\len)$ on a unit square grid, with $\tiles$ odd. Example for $\tiles = 5$%
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

Based on the preceding analysis, two cases need to be distinguished for $\tiles = \funt(\len)$ odd, depending on whether \eqref{eq: probmax len odd2} is satisfied or not; and similarly two cases exist for $\tiles = \funt(\len)$ even, depending on whether \eqref{eq: probmax len even2} is satisfied or not. These are the four cases in \eqref{eq: probmax}.

>>>>

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing than \mid
\newcommand{\diff}{\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
%\newcommand{\funca}{\psi} % function that gives average number of grid line crossings for real-valued length
\newcommand{\probmax}{\rho} % function that gives the probability of achieving the maximum number of touched tiles in the square case for real-valued length
\newcommand{\ras}{\sigma}
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genfun}{q}
\newcommand{\genvar}{s}
\newcommand{\roundterm}{r}
\newcommand{\mss}{M}
\newcommand{\nextrat}{K}
\newcommand{\funcirc}{\Omega}
\newcommand{\funcircesc}{\bar \Omega}

@@

\begin{abstract}
Consider a line segment placed on a two-dimensional grid of rectangular tiles. This paper addresses the relationship between the length of the segment and the number of tiles it visits (i.e.~has intersection with). The square grid is also considered explicitly, as some of the specific problems studied are more tractable in that particular case. The segment position and orientation can be modelled as either deterministic or random. In the deterministic setting, the maximum possible number of visited tiles is characterized for a given length, and conversely, the infimum segment length needed to visit a desired number of tiles is analyzed. In the random setting, the average number of visited tiles and the probability of visiting the maximum number of tiles on a square grid are studied as a function of segment length. These questions are related to Buffon's needle problem and its extension by Laplace.

@@

As an example of question~\ref{prob: determ dir}, consider $a=1.35$, $b=1$. A segment of unit length can be placed as shown in Figure~\ref{fig: examples} (left) to make it visit $3$ tiles. In fact, this is the maximum number for $\len=1$. The figure also illustrates that the solution for length $2.4$ is $5$ (center), and for $4.7$ it is $8$ (right).

@@

The \emph{inverse} problem~\ref{prob: determ inv} is, given $\tiles \in \mathbb N$, to determine the infimum length of all segments that visit at least $\tiles$ tiles. If the length can take any real value the infimum is not a minimum, because given any segment it can be shortened by some small amount without changing the number of visited tiles. This is a consequence of the interior of each tile being an open set.

@@

To address the remaining two questions, the notion of a \emph{random} segment of a given length needs to be precisely defined. This is done as follows. By symmetry, one endpoint of the segment can be assumed to lie in a fixed, reference tile. The position of this endpoint is \emph{uniformly} distributed on the tile. The segment orientation has a \emph{uniform} distribution on $[0,2\pi)$, and is \emph{independent} of the endpoint position. Solving the problem~\ref{prob: rand ave} of how many tiles the segment visits on average also answers, as will be seen, the inverse question (segment length to visit a given number of tiles on average). A natural, related problem~\ref{prob: rand prob} is with what probability the segment visits the maximum number of tiles.

@@

The Buffon-Laplace needle problem \cite[section~1.1]{Mathai99} considers a needle randomly placed on a grid of rectangular tiles of width $a$ and height $b$. The number of visited tiles equals one plus the number of crossings almost surely. The probability of the needle staying within a single tile is computed in \cite{Arnow94} for the case where $\len < \min\{a,b\}$. One of the problems considered in this paper, as mentioned above, is the complementary question~\ref{prob: rand prob} of the probability that the needle visits the maximum number of tiles possible for its length.

@@

For arbitrary $a, b \in \mathbb R^+$, the functions $\funt$ and $\funl$ can be computed using an iterative procedure, which exploits the fact that the pairs $(i_3,j_3), (i_4,j_4), \ldots$ of the set $\mss$ are sorted by increasing $i+j-1$, and also by increasing $(i-2)^2 a^2 + (j-2)^2 b^2$. Namely, for $\funt$ the following method yields the solution: generate successive pairs to find the last one, $(i_\tiles,j_\tiles)$, that satisfies \eqref{eq: len > sqrt}; then $\funt(\len) = \tiles$. For $\funl$ the analogous method gives a direct formula. In addition, it is possible to obtain a direct formula also for $\funt$ using a different approach. These formulas are given in Theorems~\ref{theo: funt, form} and \ref{theo: funl, form}.

@@

Theorems~\ref{theo: funt, form} and \ref{theo: funl, form} not only give the solutions $\funt(\len)$ and $\funl(\tiles)$ to the first two questions posed in \S\ref{part: intro}; they also provide a way to actually position a segment of length $\len$ or slightly greater than $\funl(\tiles)$, respectively, so that it visits $\funt(\len)$ or $\tiles$ tiles. Namely, for $\isoli$, $\jsoli$ computed as in the corresponding theorem, the segment should have its endpoints in the interiors of two tiles shifted $\isoli-1$ steps horizontally and $\jsoli-1$ steps vertically with respect to each other, with the exact position and orientation of the segment adjusted to avoid any grid points.

@@

For $0 \leq \genvar \leq 1$, let
\begin{equation}
\label{eq: f}
f(\genvar) = \int_0^\genvar \arccos z \, \diff z = \genvar \arccos \genvar - \sqrt{1-\genvar^2} + 1.
\end{equation}

@@

In the first two cases above, the probability that $x_2 > a(n-1)$, conditioned on $x_1$, is the length of the arc to the right of the line $x=a(n-1)$ divided by $2\pi\len$, that is,
\begin{equation}
\label{eq: Pr x_2 > a(n-1) cond}
\Pr[x_2 > a(n-1) \cond x_1] = \frac 1 \pi \arccos \frac{a(n-1)-x_1}{\len}.
\end{equation}
In the first case $x_1$ has a uniform distribution on $[0,a)$, and $\Pr[x_2 > a(n-1)]$ is easily obtained from \eqref{eq: Pr x_2 > a(n-1) cond} as
\begin{equation}
\label{eq: Pr x_2 > a(n-1), first case}
\begin{split}
\Pr[x_2 > a(n-1)] &= \E[\Pr[x_2 > a(n-1) \cond x_1]] \\
&= \frac 1 {\pi a} \int_0^a\arccos \frac{a(n-1)-x_1}{\len} \, \diff x_1 \\
&= \frac{\len}{\pi a}\,\, \left( f\left( \frac{a(n-1)}{\len} \right) - f\left( \frac{a(n-2)}{\len} \right) \right),
\end{split}
\end{equation}
where the function $f$ is defined in \eqref{eq: f}. Substituting into \eqref{eq: Pr i = 2 Pr} yields the result in \eqref{eq: Pr i geq n}, second line.

@@

For $u, v \in \mathbb N$, $\genvar \in \mathbb R^+$, let
\begin{equation}
\label{eq: g def}
g(\genvar, u, v) = \frac 2 \pi \int_{\arcsin \frac v {\genvar}}^{\arccos \frac u {\genvar}} \left( \genvar \cos \theta - u \right) \left( \genvar \sin \theta - v \right) \, \diff \theta.
\end{equation}
Computing this integral and using the identities $\cos^2 \alpha = 1-\sin^2 \alpha$ and $\cos(2\alpha) = 2\cos^2\alpha-1 = 1 - 2\sin^2\alpha$,
\begin{equation}
\label{eq: g fin}
\begin{split}
g(\genvar, u, v) &= \frac 1 {\pi} \biggl(
2\left(\arccos\frac{u}{\genvar}-\arcsin\frac{v}{\genvar}\right) u v + \genvar^2 + u^2 + v^2 \\
& \quad - 2 u \sqrt{\genvar^2-v^2} - 2 v \sqrt{\genvar^2-u^2} \biggr).
\end{split}
\end{equation}
Taking into account that $\arcsin x = \pi/2 - \arccos x$, it follows from \eqref{eq: g fin} that $g(\genvar, u, v) = g(\genvar, v, u)$.

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles is
\begin{equation}
\label{eq: probmax impl}
\probmax(\len) = \sum_{\substack{i,j \geq 2,\ % 
i+j-1=\tiles,\\
(i-2)^2+(j-2)^2<\len^2
}} g\left(\len, i-2, j-2 \right),
\end{equation}
or equivalently, defining $\nextrat = \sqrt{\len^2/2-(\tiles-3)^2/4}$,
\begin{equation}
\label{eq: probmax expl}
\probmax(\len) =
\begin{cases}
\displaystyle
g\left(\len, \frac{\tiles-3} 2, \frac{\tiles-3} 2 \right) +
2 \sum_{k=1}^{\lceil\nextrat-1 \rceil} g\left(\len, \frac{\tiles-3} 2 - k, \frac{\tiles-3} 2 + k \right) &\text{for $\tiles$ odd} \\[6mm] % *!*
\displaystyle
2 g\left(\len, \frac{\tiles-4} 2, \frac{\tiles-2} 2 \right) +
2 \sum_{k=1}^{\lceil\nextrat-3/2 \rceil} g\left(\len, \frac{\tiles-4} 2 - k, \frac{\tiles-2} 2 + k \right) &\text{for $\tiles$ even}.
\end{cases}
\end{equation}
\end{theorem}

@@

% For $i,j \in \mathbb Z$, let the (square) tile with lower-left coordinates $(i,j)$ be defined as $[i,i+1) \times [j,j+1)$, i.e.~including its lower and left edges.

Without loss of generality, the first endpoint of the segment can be assumed to be contained in the tile with lower-left corner $(0,0)$, which will be called the reference tile.

Consider $\len \in \mathbb R^+$, and let $\tiles = \funt(\len)$. By Proposition~\ref{prop: i+j-1}, the segment visits $\tiles$ tiles almost surely if the dimensions $i$, $j$ of its discrete bounding rectangle satisfy $i+j-1=\tiles$. In addition, a segment that visits the maximum number of tiles always has $i, j \geq 2$, because if $i=1$ the segment can always be shifted and possibly tilted to cross a vertical grid line, thus increasing $i$ to $2$, without reducing $j$. The reasoning for $j=1$ is analogous. Thus for $i, j \geq 2$, $i+j-1=\tiles$ let $\probmax_{i,j}(\len)$ denote the probability that the segment has a discrete bounding rectangle with dimensions $i, j$. Then
\begin{equation}
\label{eq: probmax len probmax i,j len}
\probmax(\len) =  \sum_{i,j \geq 2,\ i+j-1=\tiles} \probmax_{i,j}(\len).
\end{equation}

For any $i, j \geq 2$ with $i+j-1=\tiles$, the segment has a discrete bounding rectangle with dimensions $i, j$ if and only if its second endpoint is in the tile with lower-left corner $(i-1,j-1)$ or in one of the other three symmetrical tiles with respect to the lines $x=1/2$ and $y=1/2$, i.e.~those with lower-left corners $(i-1,1-j)$, $(1-i,1-j)$ and $(1-i,1-j)$. An example is shown in Figure~\ref{fig: probmax_symmetric_tiles}. Consequently, $\probmax_{i,j}(\len)$ can be obtained by computing the probability that the second endpoint is in the tile with lower-left corner $(i-1,j-1)$ and then multiplying by $4$.

\begin{figure}
\centering%
\includegraphics[width=.5\textwidth]{probmax_symmetric_tiles_}%
\caption{Symmetric tiles for the computation of $\probmax_{i,j}(\len)$ on a unit square grid. Example for $i=3$, $j=4$%
}%
\label{fig: probmax_symmetric_tiles}%
\end{figure}%

By Proposition~\ref{prop: len ineq i j}, the tile with lower-left corner $(i-1,j-1)$ can be visited only if
\begin{equation}
\label{eq: i-2 j-2 len}
(i-2)^2+(j-2)<\len^2.
\end{equation}
Thus $\probmax_{i,j}(\len)=0$ if this restriction does not hold. In the following it will be assumed that it does.

For a fixed orientation $\theta$, the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ if and only if shifting the segment up and to the right so that its first endpoint is at $(1,1)$ results in the second endpoint still being in the tile with lower-left corner $(i-1,j-1)$, as shown in Figure~\ref{fig: probmax_funt_odd}. It cannot be the case that the segment ``overshoots'' past this tile, because that would yield a number of visited tiles exceeding $\tiles$, which is the maximum. In other words, only one corner of the tile can be contained within the arc in the figure, namely its lower-left corner $(i-1,j-1)$.

\begin{figure}
\centering
\includegraphics[width=.56\textwidth]{probmax_funt_odd}%
\caption{Reachability from the reference tile for a given $\tiles = \funt(\len)$ on a unit square grid. Example for $\tiles = 7$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

There is a range of values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For each $\theta$ in the allowed range, the valid positions for the second endpoint of the segment are in a rectangle contained in the tile with lower-left corner $(i-1,j-1)$. These are the positions that cause the first endpoint to be in the reference tile. The rectangle associated with an example $\theta$ is shown in Figure~\ref{fig: probmax_funt_odd}, shaded. The area of this rectangle is the probability $\probmax_{i, j}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. As seen in the figure, this area is
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{i, j}(\len, \theta) = \left(\len\cos\theta - i+2\right) \left(\len\sin\theta - j + 2\right).
\end{equation}
The range of allowed values for $\theta$, as can also be deduced from the figure, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{j-2}{\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{i-2}{\len}.
\end{align}
For $\theta$ outside of this range the conditional probability $\probmax_{i, j}(\len, \theta)$ is $0$. Averaging $\probmax_{i, j}(\len, \theta)$ over $\theta$ gives the probability that the segment visits this tile, and then multiplying by $4$ yields $\probmax_{i.j}(\len)$:
\begin{equation}
\begin{split}
\label{eq: probmax: cond on theta}
\probmax_{i,j}(\len) &= 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{i, j}(\len, \theta)}{2\pi} \, \diff \theta \\
&= \frac 2 {\pi} \int_{\arcsin\frac{j-2}{\len}}^{\arccos\frac{i-2}{\len}} \left( \len\cos\theta - i + 2 \right)  \left( \len\sin\theta - j + 2 \right) \, \diff \theta
= g( \len, i-2, j-2).
\end{split}
\end{equation}

The above arguments show that $\probmax_{i,j}(\len)$ is given by \eqref{eq: probmax: cond on theta} if the condition \eqref{eq: i-2 j-2 len} holds, and equals $0$ otherwise. Using this into \eqref{eq: probmax len probmax i,j len} yields \eqref{eq: probmax impl} .

Given $\tiles$, the range of lengths $\len$ such that $\tiles = \funt(\len)$ is either \eqref{eq: square, odd max tiles, lengths} or \eqref{eq: square, even max tiles, lengths}. For $\len$ in this range, the number of tiles with lower-left corners $(i-1,j-1)$, $i,j \geq 2$, $i+j-1=\tiles$ that can be reached, i.e.~the number of terms in \eqref{eq: probmax impl}, depends on $\len$. If $\tiles$ is odd the minimum number is $1$, corresponding to a tile with lower-left corner on the line $x=y$, which gives the first term in the first expression in \eqref{eq: probmax expl}. For $\tiles$ even the minimum number is $2$, corresponding to two symmetrical tiles whose $x$ and $y$ coordinates differ by $1$, which give the first term in the second expression. Either for $\tiles$ odd or even, additional tiles may become reachable as $\len$ grows (within the range allowed by $\tiles$), and these always occur in symmetrical pairs. These tiles (if any) correspond to the terms in the (possibly empty) sum indexed by $k$ in either expression in \eqref{eq: probmax expl}.

Figure~\ref{fig: probmax_terms} contains examples for several values of $\tiles$, with a few values of $\len$ for each $\tiles$. Each length is represented by an arc with radius $\len$ centered at $(1,1)$. For a given $\tiles$, the filled circular markers are the lower-left corners of the tiles that are always reachable, whereas the empty circles correspond to tiles whose reachability depends on $\len$.
% For example, with $\tiles=11$ and the greatest $\len$ shown in the figure the sum over $k$ in \eqref{eq: probmax expl} has $2$ terms, corresponding to the $2$ pairs of empty markers on the diagonal line that are enclosed by the arc.

\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{probmax_terms}%
\caption{Tiles with lower-left corners $(i-1,j-1)$, $i+j-1=\tiles=\funt(\len)$ that can be reached from the reference tile on a unit square grid. Examples for several values of $\tiles$ and $\len$.
}%
\label{fig: probmax_terms}%
\end{figure}%
